---
title: "Data Wrangling in R, Lesson 1 Challenge"
author: ""
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction
This is the challenge notebook for the first data wrangling lesson. Test the skills that you picked up today by cleaning hurricane data from `dplyr`. We've filled in the dependencies section - the rest is up to you!

## Dependencies
This notebook requires two packages from the `tidyverse` as well as two additional packages:

```{r load-packages}
# tidyverse packages
library(dplyr)     # data wrangling
library(readr)     # read and write csv files

# data wrangling
library(janitor)   # data wrangling

# manage file paths
library(here)      # manage file paths
```

## Cleaning Data

In a single pipeline, you should:

1. Read the data from `data/storms_messy.csv` into `R`
2. Clean all of the very messy variable names
3. Remove the latitude and longitude variables altoghether
4. Further clean up the names of the storm name (there is a typo!), saffir simpson category, status, wind, and pressure variables

```{r clean-data}

```
